An AI is Haunting Campus^[Karl Marx wrote about a \"spectre...haunting Europe--the spectre of communism\"; GenAI looks awfully like the spectre of unregulated surveillance capitalism.]: What Role for the Human in Higher Education?

### 2025-01-25 paper/paper.qmd
Revisions for submission to Critical AI

On November 30, 2022, OpenAI released what became its flagship product, [ChatGPT](https://en.wikipedia.org/wiki/ChatGPT), a chatbot based on [Large Language Models](https://en.wikipedia.org/wiki/Large_language_model) (LLMs) designed and trained to mimic natural language. The launch of the chatbot was an impressive feat of marketing [@cowen_chatgpt_2023], and it took less than two weeks for the hype to reach stratospheric levels [@roose_brilliance_2023; @lock_what_2022]. Soon after, seemingly everyone wanted in on the Generative AI (GenAI) craze^[The pattern of cyclical hypes around AI even has a name: [AI winters](https://en.wikipedia.org/wiki/AI_winter); the current one, with chatbots playing the loudest horn atop the bandwagon, is just the most recent cycle of hype [see @siegel_ai_2023].].

According to some, AI technologies in education "offer new tools and applications that have the potential to transform traditional teaching and learning methods" [@adiguzel_revolutionizing_2023, p. 1], and have "the potential to fundamentally change the classroom experience and the knowledge and skills outcomes of our students" [@dwivedi_opinion_2023, p.20]. Furthermore, the technology "is here to stay...[t]herefore, teachers and students must develop the specific digital competencies needed to use such tools in ways that are pedagogically beneficial and ethical" [@kohnke_chatgpt_2023, p. 546].

AI was quickly painted as nothing short of revolutionary: an _innovative_ and _exciting_ technology, capable of _enhancing_ research and education, and that moreover can lead to _efficiencies_^["Efficiencies", to be clear, is the day's euphemism for cutting costs [@oconnell_let_2012, p. 64]. Cost-cutting is usually implemented by finding "redundancies" rather than finding better things for people to do.]. Institutions of higher education were compelled to get in on the act, forming high-level advisory committees and drafting guidelines to guide the university community leverage AI in principled ways [e.g., @office_of_teaching_and_learning_provisional_2023; @mcmaster_university_provisional_2024; @trent_university_generative_2024; @university_of_toronto_ai_2024].

For perspective, hammers are powerful tools, and so are thousands of other instruments found at universities. And yet not every single one commands a high-level university committee to advice on their use. Just what is it about AIs that requires pan-university consultations on how to use them? To understand why practically every institution of higher education has decided to spend considerable resources studying the adoption and responsible use of AI, one needs to understand what the technology claims to do that sets it apart from other tools--and then, examine what it actually does.

In doing this, we must remain alert to the fact that technologies are never politically neutral, and two questions must be asked when considering their adoption:

@@

This essay aims to provide some clarification around 1) claims about the potential of GenAIs in education; and 2) what GenAIs actually do--within 3) the context of a principled organization. I conclude that the promise of generative AI has been overstated, that is is unlikely to deliver, and that in any case the technology, as sold by the tech industry, is antithetical to many of the values that universities claim to hold dear.

# The university's mission {#the-mission}

Universities present themselves as institutions that serve "the social, cultural, and economic needs of [their] community and [their] society" [@office_of_the_president_mcmaster_university_mission_2024]; that strive "to make valued and socially responsible contributions to our local communities, to Canada, and to the world" [@trent_university_vision_2024]; and that "serve society and...enhance the quality of life through scholarship" [@university_of_guelph_mission_2024].

This goal is achieved through "the discovery, communication and preservation of knowledge" [@office_of_the_president_mcmaster_university_mission_2024]; a dedication "to fostering an academic community in which the learning and scholarship of every member may flourish" [@the_office_of_the_governing_council_secretariat_institutional_1992]; by "[encouraging] and [celebrating] excellence and innovation in teaching, learning, research and student development" [@trent_university_vision_2024]; with a determination "to put the learner at the centre of all [the university] does" [@university_of_guelph_mission_2024].

Unprincipled organizations do not need to question their choices of technology--as long as the technology advances whatever its goal is at any particular moment. But universities aim to achieve their mission while adhering to "integrity, quality, inclusiveness and teamwork" [@office_of_the_president_mcmaster_university_mission_2024]; with "vigilant protection for individual human rights, and a resolute commitment to the principles of equal opportunity, equity and justice" [@the_office_of_the_governing_council_secretariat_institutional_1992]; and  as they "foster sustainability, in its environmental, social and economic dimensions" [@university_of_guelph_mission_2024].

@@

A reason why many organizations have spent substantial resources considering the adoption of AIs is that they have been presented, quite deliberately, as _everything tools_: tools capable of doing many things a human can do, but faster and supposedly cheaper. While hammers are understood to have a limited range of uses that mostly have to do with blunt force, the release of GenAIs to the masses was accompanied by innumerable promises--quite a few of them made by entities and individuals that had a stake in the adoption of the technology^[Ranging from "AI is inevitable, and I will sell it to you" to "AI is inevitable, and for a consulting fee I will tell you how to level up with it".].

The public was promised that GenAI would "accelerate sustainability" [@nakagawa_accelerating_2023].

Also, that LLMs would transform sectors as diverse as health care, finance, human resources, insurance, "and beyond" [@ibm_what_2023]. AI-powered self-driving cars would reduce accidents, car ownership, pollution and noise, and would automate logistics and make parking tickets obsolete [@joshi_5_2022; @garsten_what_2024]. In mental health care, chatbots could complement human therapists by using patient data to analyze behavioral patterns to then mimic practitioner questions [@silva_4_2023]. AIs would revolutionize dating by helping "people connect and find potential partners" [@khalatian_matchmaking_2023].

The list of promises about what AI _can_ do has grown to encompass pretty much everything, from saving the planet from ourselves to micromanaging the [bowel movements of our pets](https://finance.yahoo.com/news/next-generation-ai-powered-cat-140000795.html).

```{r ai-litter-box, out.width="70%", fig.cap="\\label{fig:ai-bug-fix}Revolutionizing pet health"}
#| label: fig-revolutionizing-pet-care

knitr::include_graphics(glue::glue(here::here(), "/images/ai-litter-box.png"))
```

@@

@power_3_2024, presents three innovative uses for AI in higher education. First, AI could take over manual tasks and improve the budget by releasing administrative staff^[Probably from working at the university]. AI could evaluate applications for admissions, including assessing traits like the "grit" and "empathy" of applicants. AI could also personalize the experience of consumers by predicting their final grade with 60-70\% accuracy, and by providing corrective and instructive feedback.

In another instance, @hie_how_2023 argue that AIs will likely be the primary way to access information in the future. For this reason, institutions must equip students with the skills like _prompt engineering_, the ability to craft effective queries that prompt a GenAI to provide the most useful outputs. Universities will use AIs to improve learning and challenge the critical thinking of students. Faculty will also benefit from AI: it will be used to evaluate their students, and will handle repetitive tasks like writing course syllabi.

@hie_how_2023 warn that "AI is here to stay"^[Writing in 2022, Mr. Hié predicted that the metaverse would be a [revolution in higher education](https://www.linkedin.com/pulse/metaverse-dedicated-pedagogy-revolution-higher-education-anthony-hi%C3%A9). This prediction aged poorly: on February, 2023, Microsoft [disbanded](https://www.theinformation.com/articles/microsoft-kills-its-industrial-metaverse-team-after-4-months) its Industrial Metaverse Core Team scantly four months after launching it. Shortly after that, Meta [slashed its own metaverse team](https://www.reuters.com/technology/meta-lay-off-employees-metaverse-silicon-unit-wednesday-2023-10-03/) in October, 2023.], and that "[faculty] might not be able to stop the AI trend from growing" so they might as well get on with the program and try to shape it as best they can.

@@

Is it really possible that AIs can do all this?

@@

Discussions of AI as an _everything tool_ quickly devolve into a [Gish Gallop](https://en.wikipedia.org/wiki/Gish_gallop), a rhetorical device where a party makes an overwhelming number of arguments with little regard to their accuracy or strength. To be sure, AIs can do _some_ things well, but critics are forced to respond to every single propounded use with evidence, which usually is much more time consuming than making the unsubstantiated and/or misleading claims of the gallop itself. Proponents of AI can always admit that, ok, AIs do not actually do _that one thing_, but what about the myriad of other things that people say they really do? A second discursive device deployed by proponents of AI is an informal fallacy known as [The False Dilemma](https://en.wikipedia.org/wiki/False_dilemma): adopt AI or fail to level up.

In what follows, it is not my intention to confront every single possible use of AI in teaching and learning, since doing so plays into the intent of the Gish Gallop. As well, I reject the dilemma posed by the purveyors of hype, and I remain convinced that there are multiple ways to excel at teaching and learning that do not require AI, let alone the compromises demanded by AI.

@@

GenAIs are _deep learning models_ [@ibm_what_2021], so-called because they are neural networks with _depth_, meaning _many_ hidden layers. As an example, Large Language Models are models built with multiple layers of neural networks trained on _very_ large amounts of textual data [@ibm_what_2023].

@@

A regression model takes data as inputs, typically a "dependent" variable $y$ (observations of an outcome of interest) and "independent" variables $x$ (observations of things thought to correlate with the dependent variable). With these inputs, the model finds a conditional mean that becomes the output of the model, or $\hat{y}$ (i.e., a mean value conditional on the values of the dependent variables). The process of finding that conditional mean is to satisfy some criterion (e.g., that the conditional mean minimizes the error of the model--the distance between $y$ and $\hat{y}$).

@fig-bivariate-model is an example of the simplest regression model with only one independent variable (i.e., a bivariate model). The plot shows the pairs of $y$ and $x$ values, and the line is the model, i.e., the conditional mean $\hat{y}$. This model needs only two parameters: an intercept (the conditianl mean $\hat{y}$ when $x = 0$), and the slope of the line, in other words, the rate of change of $\hat{y}$ with respect to $x$. Each additional independent variable uses one additional parameter to describe the slope of the model with respect to that variable.

@@

Neural networks do something similar but using a more flexible net-like architecture. For instance, @fig-nn1 shows a relatively simple neural network trained using using the same data. This network is not "deep", as it has only one "hidden layer" comprised of two neurons between the input and the output layers. In comparison to the linear regression and its two parameters, this neural network uses seven parameters. Another key difference between a linear regression model and a neural network is that the latter requires an initial set of parameters, which are often drawn at random, thus making the results of the model contingent on the starting conditions of the training process.
```{r}
# Given the dataframe grid.1, let us now train a neural network. We need to first define a formula that specifies the dependent variable and the independent variables.
f <- as.formula("y ~ x")
```

@@

@fig-data-vs-preds-02 illustrates the predictions of the neural network in @fig-nn1, where we can see that the model is not limited to a linear response. This flexibility by the neural network is achieved by being less parsimonious (it uses more parameters).

@@

More complex neural networks can be designed. @fig-nn2 shows a network deeper than the previous one, with two hidden layers, each consisting of five neurons. This model uses forty-six parameters. 
```{r train-deeper-neural-network}
# Train a neural network using only two neurons in one hidden layer. The argument `linear.output` is set to TRUE, so that the activation function is not applied to the output node. The default activation function is the logistic function.
set.seed(242)
nn2 <- neuralnet(f, 
                data = df, 
                hidden = c(5, 5), 
                linear.output = T)
```

@@

There is a limit as to how flexible a model can be, and this is true of linear regression and of neural networks. Each additional parameter makes the model more flexible, but as a rule we cannot use more parameters than the number of data points that exist in the dataset used to train the model. Roughly, every additional parameter needs one additional datum to train on, and when there are as many parameters as data inputs the model becomes overfit and will predict each value of the independent variable in the training dataset perfectly. The model will fail to train if there are more parameters than data points.

In a neural network trained to work with natural language, the "predictors" are the words that come before and after a certain word. Several approaches exist to _embed_ a bag of words, which means using large text corpora to calculate the frequency of co-occurrence of words. These frequencies are then converted to numerical vectors that represent the position of a word in multidimensional space. These "coordinates" in the space of the vocabulary can then be used to calculate the "distances" between words. A recent breakthrough in the field of natural language processing is the development of _transformers_, a network architecture architecture that updates the numeric values of a word depending on the words that come before and after in a prompt.

Here are some key ideas to keep in mind when thinking about AIs:

- AIs are neural networks, which is to say, regression models.
- A regression model generates outputs that are regressions to the mean (i.e., statistically probable outputs).
- AIs are _very_ flexible non-linear regression models.
- Their flexibility comes from using "deep" architectures, which means many layers with many parameters^[According to reports [see @griffith_gpt_4_2023], GPT-1 had 117 million parameters, GPT-2 1.5 billion, and GPT-3 175 billion parameters. OpenAI no longer reports the number of parameters of its newer models, but in the latest version this is estimated to be in the trillions.]. The large number of parameters makes these models _extremely_ data hungry. 
- AIs are trained using an initial set of random values as parameters. The same training dataset can result in different models due to the randomness in the initial conditions.
- The flexibility of AIs means that outputs will vary seemingly at random with relatively small changes in the prompt. 
- Also, importantly, since regression models aim for the mean (i.e., a statistically probable output), they tend to be bad at _extrapolation_ (or thinking outside the box, if AIs could think). The further away an input (i.e., prompt) is from the space defined by the training dataset, the worse the output will tend to be^[Which is possibly why so-called hallucinations became more subtle as bigger datasets were used to train AIs. The increasingly large datasets essentially expanded the space where the model can interpolate.].

@@

Some of the things that universities hope to achieve with AI include writing text (e.g. reference letters, grant applications, job descriptions, report summaries), conducting qualitative and quantitative data analysis, translation, captioning, providing multiple formats for learning and engagement, and for personalized support. Anecdotally, students report wanting to use generative AI not to complete assignments for them, but to help enhance learning in their courses by having concepts explained differently, by working on practice questions or by getting immediate feedback on drafted work. If we take a student-centered perspective, we might begin by asking about some of the things that AIs purportedly do for students.

## What do GenAIs do for students?{#what-do-ais-do-for-students}

- Students want concepts explained to them differently.

@@

- Practice questions

@@

- Immediate feedback on drafted work

@@

- Creation of innovative learning materials

@@

- Automated grading

@@

- Writing reference letters

@@

- Qualitative and quantitative data analysis

@@

An error like this is easy to spot. But we seldom use computers to do calculations that can be done by hand, and checking that an AI-generated response is not wrong is as time consuming as doing the analysis without an AI in the first place [see @yuan_how_2023].

From a different perspective, even the average output of an AI should be treated with caution. As noted [above](#really-what-do-they-do), the results of a neural network depend on the weights selected at random at the start of a training process. Change those weights, the results will change. As well, even small variations in the prompt can be impactful in models with billions or trillions of parameters. In statistical terms, any output is but one _realization_ of a random process. For the results to be useful, we should calculate the average of a large number of responses. Given the cost of training and querying a LLM, this is prohibitively expensive.

But what about a task that involves logic instead of arithmetic? Research by @nezhurina_alice_2024 demonstrates that relatively simple natural language puzzles that are easily solved by humans, often result in confidently expressed errors when used to prompt an AI. Wrong answers, furthermore, are frequently backed up by confabulations, that is, nonsense expressed in reasoning-like terms. The onnections between words that LLMs deliver are but probabilistic word salads: @xu_hallucination_2024 use a formal approach to show that LLMs are inherently incapable of mapping outputs to a ground truth, and thus their results will only ever resemble a ground truth purely by chance. The outputs of a chatbot only sound plausible thanks to the vast amount of data points and parameters that the models use. But if the objective of qualitative research is to identify interesting aspects of the human experience, it is certainly not the case that they will be found in a regression to the mean.

- Identify early signs of learning difficulties.

An AI should not be used to identify signs of learning difficulties for the same reasons that AIs cannot be trusted to identify text written by an AI. The false positive and negative rates are out of proportion compared to the benefits the practice might offer. A vigilant instructor will be aware of learning difficulties among their students. But were the process to become automated, who would decide when an intervention was warranted, and who qualified for it? Would the intervention be offered to everyone flagged by the system, including the false positives? And, could those denied an intervention complain? Importantly, who would fund an intervention, and who would deliver it? Another AI?

- Personalized learning

Anyone receiving "personalized learning" from an AI can be sure that they will be treated as the machine-generated average of their class, and not as a person.

## What do GenAIs do to students? {#what-can-ais-do-to-students}

The preceding section illustrates how numerous tasks that AIs can notionally do in higher education are in fact poor replacements for a human-centered education. But what about the things that GenAIs can do _to_ students?

@@

Reliance on AI can short-circuit the learning process. Statements like "in the very near future AIs will likely be the primary way we access knowledge" [@hie_how_2023] convey the idea that prompt engineering will become an essential skill; perhaps even the skill to end all skills^[Prompt engineering is a pompous name for a task that more closely resembles reading the entrails of a Large Language Model than actual engineering. It is a task that cannot possibly be performed in a systematic way, and that offers no guarantees that it will work in the same way again, especially after a model is updated. Calling the activity of writing prompts "engineering" is a blatant attempt to steal the competence and reputation of actual engineering.].

It is important to note that, even if prompt "engineering" offered some value, the skill is tied to a singular technology and offers little transferability potential. @mollick_instructors_2024 demonstrate how prompt "engineering" is not even transferable _between_ different LLMs (see p. 5 and p. 36). Good skills transfer between contexts. Learning how to do an effective bibliographical query transfers between libraries. Learning to identify relevant content works while reading a novel, a book of history, or a text of econometrics. Writing, the slow and demanding process of converting thoughts into words, is essential to create a style of communication that is unique to each person. It might be argued, in fact, that writing _is_ learning [@klein_trends_2016].

In a sense, an AI is like a crutch. However, what the public is being promised is not a crutch: it is an All-Terrain Vehicle, but one that the user cannot hope to fully master due to its blackbox nature--in the sense that 1) neural networks are blackboxes; and 2) the development of AIs itself is a blackbox wherein things can more easily change based on the interests of their corporate owners than in response to the interests of their relatively small customers^[And this refers not only to individuals whose skills become yoked to ChatGPT and MidJourney. As an example, in 2022 McMaster University operated on revenues of less than [\$843 million](https://financial-affairs.mcmaster.ca/app/uploads/2023/06/2023-24-Consolidated-Budget-FINAL.pdf) (CAD). Microsoft, in contrast, operated on revenues of approximately [\$198,000 millions](https://www.microsoft.com/investor/reports/ar22/) (USD) in 2022. To put things in perspective (and using the average exchange rate of 2022), if McMaster used _all_ of its 2022 revenue to pay Microsoft, that would leave McMaster with zero dollars and still account for less than _one third of one percent point_ of Microsoft's revenue. A mid-sized Canadian university is like an ant scuttling under the economic boot of Microsoft.].

@@

Besides a few methodological developments, improvement in the performance of AIs has mostly relied on using bigger datasets. This has led to a rush, as developers try to train their models with bigger datasets^[A rush that feels as frenetic as any gold rush in the past. However, while the gold rushes of the 19th century were relatively democratic due to their low entry costs, the current data rush has enormous entry costs and is largely dominated by large corporations and venture capital.]. But, despite the large number of text items and images available in digital form, there is still only a finite amount of data to be had.

As of this writing, the few big players that can afford to grab the data, are mostly done with all the data grabbing, sometimes by morally dubious, and at times by possibly illegal means [@clark_ex-amazon_2024]. And so, as the fountain of human-generated content has been reduced to a drip by a handful of corporations developing AIs, there has been a turn towards what might be the last untapped drops of information left on the internet.

>>>>

On November 30, 2022, OpenAI released what became its flagship product, [ChatGPT](https://en.wikipedia.org/wiki/ChatGPT), a chatbot based on [Large Language Models](https://en.wikipedia.org/wiki/Large_language_model) (LLMs) designed and trained to mimic natural language. The launch of the chatbot was an impressive feat of marketing [@cowen_chatgpt_2023], and it took less than two weeks for the hype to reach stratospheric levels [@roose_brilliance_2023; @lock_what_2022]. Soon after, everyone wanted in on the Generative AI (GenAI) craze^[The pattern of cyclical hypes around AI even has a name: [AI winters](https://en.wikipedia.org/wiki/AI_winter); the current one, with chatbots leading the parade, is just the most recent cycle of hype [see @siegel_ai_2023].].

AI technologies in education "offer new tools and applications that have the potential to transform traditional teaching and learning methods" [@adiguzel_revolutionizing_2023, p. 1], and have "the potential to fundamentally change the classroom experience and the knowledge and skills outcomes of our students" [@dwivedi_opinion_2023, p.20]. Furthermore, the technology "is here to stay...[t]herefore, teachers and students must develop the specific digital competencies needed to use such tools in ways that are pedagogically beneficial and ethical" [@kohnke_chatgpt_2023, p. 546].

AI is painted as revolutionary: an _innovative_ and _exciting_ technology, capable of _enhancing_ research and education, while leading to _efficiencies_^["Efficiencies" being an euphemism for cutting costs [@oconnell_let_2012, p. 64], usually by finding "redundancies" rather than better things for people to do.]. Higher education institutions were compelled to get in on the act, forming high-level advisory committees and drafting guidelines for university communities to leverage AI [e.g., @office_of_teaching_and_learning_provisional_2023; @mcmaster_university_provisional_2024; @trent_university_generative_2024; @university_of_toronto_ai_2024].

But hammers are powerful tools, and so are thousands of other instruments found at universities. And yet not all of them get high-level university committees to advice on their use. Just what is it about AIs that requires pan-university consultations on how to use them? To understand why practically every institution of higher education has decided to spend considerable resources studying the adoption and responsible use of AI, one needs to understand what the technology claims to do that sets it apart from other tools--and then, also examine what it actually does.

In doing this, we must remain alert to the fact that technologies are never politically neutral, and two questions must be asked:

@@

This essay examines claims about the uses of GenAIs in education, a sober assessment of what they actually do, while keeping in mind the mission of nominally principled organizations. I conclude that promises about GenAI are overstated and unlikely to be met, and that current AIs are antithetical to the values that universities claim to hold dear.

# The mission {#the-mission}

Universities pride themselves of serving "the social, cultural, and economic needs of [their] community and [their] society" [@office_of_the_president_mcmaster_university_mission_2024]; of striving "to make valued and socially responsible contributions to our local communities, to Canada, and to the world" [@trent_university_vision_2024]; and of "[serving] society and...[enhancing] the quality of life through scholarship" [@university_of_guelph_mission_2024].

These goals are achieved through "the discovery, communication and preservation of knowledge" [@office_of_the_president_mcmaster_university_mission_2024]; a dedication "to fostering an academic community in which the learning and scholarship of every member may flourish" [@the_office_of_the_governing_council_secretariat_institutional_1992]; by "[encouraging] and [celebrating] excellence and innovation in teaching, learning, research and student development" [@trent_university_vision_2024]; and with a determination "to put the learner at the centre of all [the university] does" [@university_of_guelph_mission_2024].

Unprincipled organizations need not question their choices of technology--as long as the technology advances whatever their goal might be at any moment. But universities aim to achieve their mission while adhering to "integrity, quality, inclusiveness and teamwork" [@office_of_the_president_mcmaster_university_mission_2024]; with "vigilant protection for individual human rights, and a resolute commitment to the principles of equal opportunity, equity and justice" [@the_office_of_the_governing_council_secretariat_institutional_1992]; and  as they "foster sustainability, in its environmental, social and economic dimensions" [@university_of_guelph_mission_2024].

@@

One reason many organizations have spent substantial resources on AIs is that they have been presented as _everything tools_: a technology capable of doing many things humans do, but faster and supposedly cheaper. While most tools are understood to have a limited range of uses, the release of GenAIs was accompanied by innumerable promises--most of them made by entities and individuals with a stake in the adoption of the technology^[Ranging from "AI is inevitable, and I will sell it to you" to "AI is inevitable, and for a consulting fee I will tell you how to level up with it".].

The list of promises have grown to encompass pretty much everything, from saving the planet [by "accelerating sustainability"; @nakagawa_accelerating_2023] to "doubling human lifespans" within the decade [@novak_anthropic_2025]. Transforming health care, finance, human resources, insurance, "and beyond" [@ibm_what_2023] pales besides such momentous accomplishments.

@@

For example, @power_3_2024, presents three "innovative" uses for AI in higher education. First, handing over manual tasks to AIs could improve the budget by "releasing" administrative staff. AI could assess traits like the "grit" and "empathy" in admission applications. AI could also personalize the experience of consumers [i.e., students] by predicting their final grade with 60-70\% accuracy, and by providing corrective and instructive feedback.

@hie_how_2023 argue that AIs will be the primary way to access information in the future. For this reason, institutions must equip students with skills like _prompt engineering_, the ability to craft effective queries for GenAIs. Universities will use AIs to improve learning and challenge students' critical thinking. Faculty will also benefit from AI by automating student evaluation and writing course syllabi. Moreover, @hie_how_2023 warn that "AI is here to stay"^[Writing in 2022, Mr. Hié predicted that the metaverse would be a [revolution in higher education](https://www.linkedin.com/pulse/metaverse-dedicated-pedagogy-revolution-higher-education-anthony-hi%C3%A9). This prediction aged poorly: on February, 2023, Microsoft [disbanded](https://www.theinformation.com/articles/microsoft-kills-its-industrial-metaverse-team-after-4-months) its Industrial Metaverse Core Team scantly four months after launching it. Shortly after that, Meta [slashed its own metaverse team](https://www.reuters.com/technology/meta-lay-off-employees-metaverse-silicon-unit-wednesday-2023-10-03/) in October, 2023.], and that "[faculty] might not be able to stop the AI trend from growing" so they should at least try to shape it as best they can.

@@

Is it really possible for AIs to do all this?

@@

Presenting AI as an _everything tool_ is purposeful. This device makes debates around the technology quickly devolve into a [Gish Gallop](https://en.wikipedia.org/wiki/Gish_gallop), a rhetorical style where a party makes an overwhelming number of arguments with little regard to their accuracy or strength. To be sure, AIs can do _some_ things well, but critics are forced to counter every suspect promise with evidence, a more time consuming task than making unsubstantiated and/or misleading claims. Proponents of AI can always admit that AIs may not actually do _that one thing_, but what about the myriad of other things that people say they really do? A second discursive device deployed by proponents of AI is an informal fallacy known as [The False Dilemma](https://en.wikipedia.org/wiki/False_dilemma): adopt AI or fail to level up.

It is not my intention to confront every single possible use of AI in teaching and learning, since doing so plays into the intent of the Gish Gallop. As well, I reject the false dilemma, and remain convinced that there are multiple ways to excel at teaching and learning that do not require AI, let alone submitting to the compromises demanded by AI.

@@

GenAIs are _deep learning models_ [@ibm_what_2021], so-called because they are neural networks with _depth_, meaning _many_ hidden layers. Large Language Models are models built with multiple layers of neural networks trained on _very_ large amounts of textual data [@ibm_what_2023].

@@

A regression model uses as inputs a "dependent" variable $y$ (observations of an outcome of interest) and "independent" variables $x$ (observations of things thought to correlate with the dependent variable) to find a conditional mean that becomes the output of the model, or $\hat{y}$ (i.e., a mean value conditional on the values of the dependent variables). The process of finding that conditional mean is to satisfy some criterion (e.g., that the conditional mean minimizes the error of the model).

@fig-bivariate-model is an example of the simplest regression model with only one independent variable. The plot shows the pairs of $y$ and $x$ values, and the line is the model, i.e., the conditional mean $\hat{y}$. This model needs only two parameters: an intercept (the conditional mean $\hat{y}$ when $x = 0$), and the slope of the line (the rate of change of $\hat{y}$ with respect to $x$). Each additional independent variable uses one additional parameter to describe the slope of the model with respect to that variable.

@@

Neural networks do something similar but using a net-like architecture. For instance, @fig-nn1 shows a relatively simple neural network trained using using the same data as above. This network is not "deep", as it has only one "hidden layer" (the two neurons between the input and the output layers). In comparison to the linear regression and its two parameters, this neural network uses seven parameters that need to be pre-loaded with random values, which makes the results of the model contingent on the starting conditions of the training process.
```{r}
# Given the dataframe grid.1, let us now train a neural network. We need to first define a formula that specifies the dependent variable and the independent variables.
f <- as.formula("y ~ x")
```

@@

Unlike a linear regression model, the predictions of the neural network can be non-linear (@fig-data-vs-preds-02 ). This flexibility is achieved by being a less parsimonious model (it uses more parameters).

@@

More complex neural networks can be designed. @fig-nn2 shows a deeper network, with two hidden layers of five neurons each. This model uses forty-six parameters. 
```{r train-deeper-neural-network}
# Train a neural network using only two neurons in one hidden layer. The argument `linear.output` is set to TRUE, so that the activation function is not applied to the output node. The default activation function is the logistic function.
set.seed(242)
nn2 <- neuralnet(f, 
                data = df, 
                hidden = c(5, 5), 
                linear.output = T)
```

@@

There is a limit as to how flexible a regression model can be. Each additional parameter makes the model more flexible, but as a rule we cannot use more parameters than the number of data points that exist in the dataset used to train the model. Roughly, every additional parameter needs one additional datum to train on, and when there are as many parameters as data inputs the model becomes overfit and predicts each value of the independent variable in the training dataset perfectly. The model will fail to train if there are more parameters than data points.

In a neural network trained to work with natural language, the "predictors" are the words that come before and after a certain word. Several approaches exist to _embed_ a "bag" of words, which means using massive amounts of text to calculate how frequently words go together. These frequencies are then converted to numerical vectors that represent the "position" of a word in multidimensional space, and then the "distances" between words.

Here are some key concepts to keep in mind when discussing AIs:

- AIs are neural networks, which is to say, regression models.
- Regression models generate outputs that are regressions to the mean.
- AIs are _very_ flexible non-linear regression models thanks to the use of "deep" architectures, which means many layers with many parameters^[According to reports [see @griffith_gpt_4_2023], GPT-1 had 117 million parameters, GPT-2 1.5 billion, and GPT-3 175 billion parameters. OpenAI no longer reports the number of parameters of its newer models, but in the latest version this is estimated to be in the trillions.]. 
- The large number of parameters makes these models _extremely_ data hungry. 
- AIs are trained using an initial set of random values. The same training dataset can result in different models due to the randomness in the initial conditions.
- The flexibility of AIs means that outputs will vary seemingly at random with relatively small changes in the prompt. 
- Also, importantly, since regression models aim for the mean (i.e., a statistically probable output), they tend to be bad at _extrapolation_ (or thinking outside the box, if AIs could think). The further away an input (i.e., prompt) is from the space defined by the training dataset, the worse the output will tend to be^[Which is why so-called hallucinations became more subtle as bigger datasets were used to train AIs. The increasingly large datasets essentially expanded the space where models could interpolate.].
- However, when data around a topic are sparse, the regression model can output exactly the training data, because the correlation is perfect.

@@

Uses of AI in universities include writing text (e.g. reference letters, grant applications, job descriptions, report summaries), conducting qualitative and quantitative data analysis, translation, captioning, providing multiple formats for learning and engagement, and for personalized support. Anecdotally, students report wanting to use generative AI not to complete assignments for them, but to help enhance learning in their courses by having concepts explained differently, by working on practice questions or by getting immediate feedback on drafted work. In short, AIs are supposed to help both teachers and students.

### Different explanations of concepts

@@

### Practice questions

@@

### Immediate feedback on drafted work

@@

### Creation of innovative learning materials

@@

### Automated grading

@@

### Writing reference letters

@@

### Qualitative and quantitative data analysis

@@

An error like this is easy to spot. But we do not use computers to do calculations that can be done by hand, and checking the correctedness of an AI-generated response is as time consuming as doing the analysis without an AI [see @yuan_how_2023].

From a different perspective, even the average output of an AI should be treated with caution. As noted [above](#really-what-do-they-do), the results of a neural network depend on the weights selected at random at the start of a training process. Change those weights, the results will change. As well, even small variations in the prompt can be impactful in models with billions or trillions of parameters. In statistical terms, any output is but one _realization_ of a random process. For the results to be useful, we should calculate the average of a large number of responses. Given the cost of training and querying LLMs, this is prohibitively expensive.

But what about a task that involves logic instead of arithmetic? Research by @nezhurina_alice_2024 demonstrates that relatively simple natural language puzzles that are easily solved by humans often result in confidently expressed errors when used to prompt an AI. Wrong answers, furthermore, are frequently backed up by confabulations, that is, nonsense expressed in reasoning-like terms. The connections between words that LLMs deliver are but probabilistic word salads: @xu_hallucination_2024 use a formal approach to show that LLMs are inherently incapable of mapping outputs to truths, and thus their results will only ever resemble truth purely by chance. The outputs of a chatbot only sound plausible thanks to the vast amount of data points and parameters that the models use. But if the objective of qualitative research is to identify interesting aspects of the human experience, it is certainly not the case that they will be found in a regression to the mean.

### Identify early signs of learning difficulties.

An AI should not be used to identify signs of learning difficulties for the same reasons that AIs cannot be trusted to identify text written by an AI. The false positive and negative rates are disproportionate compared to the benefits. A vigilant instructor will be aware of learning difficulties among their students. But were the process to become automated, who would decide when an intervention was warranted, and who qualified for it? Would the intervention be offered to everyone flagged by the system, including the false positives? And, could those denied an intervention complain? Importantly, who would fund an intervention, and who would deliver it? Another AI?

### Personalized learning

Anyone receiving "personalized learning opportunities" from an AI can be sure that they will be treated as the machine-generated average of the dataset used to train the model, and not as a person.

## What do GenAIs do to students and teachers? {#what-can-ais-do-to-students}

The preceding section illustrates numerous AI tasks in higher education are in fact poor replacements for a human-centered education. But what about the things that GenAIs can do _to_ students?

@@

Reliance on AI short-circuits the learning process. Statements like "in the very near future AIs will likely be the primary way we access knowledge" [@hie_how_2023] imply that prompt engineering will become an essential skill, perhaps even the skill to end all skills^[Prompt engineering is a pompous name for a task that more closely resembles reading the entrails of a Large Language Model than actual engineering. It is a task that cannot be performed in a systematic way, and that offers no guarantees that it will work in the same way again, between models or versions. Calling the activity of writing prompts "engineering" is to steal the competence and reputation of actual engineering.].

It is important to note that, even if prompt "engineering" offered some value, the skill is tied to a singular technology that offers little transferability potential. @mollick_instructors_2024 demonstrate how prompt "engineering" is not even transferable _between_ different LLMs (see p. 5 and p. 36). Good skills transfer between contexts. Learning to do effective bibliographical queries transfers between libraries. Learning to identify relevant content works for reading a novel, a book of history, or a text of econometrics. Writing, the slow and demanding process of converting thoughts into words, is essential to create a style of communication that is unique to each person. It might be argued, in fact, that writing _is_ learning [@klein_trends_2016].

An AI is like a crutch, but what we are being promised is an All-Terrain Vehicle, however one that the user will never own, and cannot hope to fully master due to its blackbox nature (neural networks are blackboxes and their development is also a blackbox wherein things can more easily change based on the interests of their corporate owners than in response to the interests of their relatively small customers^[And this refers not only to individuals. As an example, in 2022 McMaster University, a major Canadian university, operated on revenues of less than [\$843 million](https://financial-affairs.mcmaster.ca/app/uploads/2023/06/2023-24-Consolidated-Budget-FINAL.pdf) (CAD). Microsoft, in contrast, operated on revenues of approximately [\$198,000 millions](https://www.microsoft.com/investor/reports/ar22/) (USD) in 2022. To put things in perspective (and using the average exchange rate of 2022), if McMaster used _all_ of its 2022 revenue to pay Microsoft, that would leave McMaster with zero dollars and still account for less than _one third of one percent point_ of Microsoft's revenue.].

@@

Improvement in the performance of AIs has relied on using bigger datasets. This has led to a rush, as developers train their models using bigger datasets^[A rush that feels as frenetic as any gold rush in the past. However, while the gold rushes of the 19th century were relatively democratic due to their low entry costs, the current data rush has enormous entry costs and is dominated by large corporations and venture capital.]. But, despite the large number of text items and images available in digital form, the supply is finite.

As of this writing, the few big players that can afford to grab the data, are mostly done with it, using morally dubious and possibly illegal means [@clark_ex-amazon_2024; @milmo_zuckerberg_2025; @smalley_linkedin_2025]. And so, as the fountain of human-generated content has been reduced to a drip, there has been a turn towards what might be the last untapped drops of information left on the internet.
